An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-14 ju-hofmann/ju-hofmann.Rmd
fulfilled the "marking text" section (hopefully)

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullets
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian W. Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
abstract: >
  In this paper, we introduce the R package cmpsR, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package bulletxtrctr. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of Squares Ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 100,
  warning = FALSE
)
```

@@

**Hamby Data Set**
<!-- ### Hamby Data Set -->

@@

The automated framework used in this paper was proposed by @aoas.
In order to obtain bullet signatures from the x3p files of bullet land engravings, we use the `R` packages `x3ptools` [@x3ptools] and `bulletxtrctr` [@bulletxtrctr].
`x3ptools` is a package to read, write, and generally, process x3p files.
The `bulletxtrctr` package implements a pipeline for extracting and comparing signatures from scans of land-engraved areas.

@@

The white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], `bulletinspectR` to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}. Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

In this example, the comparison between `sig1` and `sig2`, two signatures coming from the same source (a known-match comparison), gets a CMPS score of 17; the comparison between `sig1` and `sig3`, two signatures coming from different sources (a known non-match comparison), gets a CMPS score of 1.

@@

1.  **cut the comparison signature into consecutive, non-overlapping and equal-length basis segments:** The command `get_segs(x, len=50)` implements this step: it takes bullet signature `x` in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length `len`, which are referred to as "basis segments".

@@

    -   **Calculate the cross-correlation curve:** Calculate the cross-correlation curve between a basis segment `x` and the reference signature `y` using the function `get_ccf4(x, y, ...)` as shown in \autoref{fig:segplot2}. The position indicates the lag by which a basis segment is moved with respect to its original placement. A position is considered "good" if it results in a peak in the cross-correlation between the basis segment and the reference.

    -   **Correlation peaks:** Two strategies referred to as "multi-peak inspection" and "multi-peak inspection at different segment lengths" in @cmps are used for identifying positions of correlation peaks as candidate positions. The latter is also called "multi-segment lengths strategy". The parameter `npeaks.set` in `extract_feature_cmps(...)` determines which strategy to use and the number of candidate positions:

        -   If `npeaks.set` is an integer vector of length 1, for example `npeaks.set = 5`, the positions of the top five peaks in the cross-correlation curve are identified as candidate positions for registration.
        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi-segment lengths strategy will be used: calculate the cross-correlation function between a basis segment and the reference and identify positions of the top five peaks; double the segment length to a specified value, re-calculate the cross-correlation function, and identify three peaks; repeat this process and identify a single peak in the newly computed cross-correlation function. \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks. Note that in @cmps the segment length is doubled at each level of a basis segment, but in the present implementation users are allowed to choose the segment length at each level.
        -   `get_ccr_peaks(comp, segments, seg_outlength, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or increased segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_outlength`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the reference signature. If the multi-segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each level of the basis segment. The resulting list is called `ccr.list`.

    -   **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment levels. Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak, but we do not have this requirement in our implementation.

        -   the function `get_seg_scale(segments, nseg, out_length)` is used to obtain the (potentially increased) version of a basis segment. `segments`, which is a list containing all basis segments generated by the function `get_segs(...)` in step 1, and `nseg` are used to determine the basis segment to be increased. `out_length` specifies the length of the output segment.
        -   `get_ccp(ccr.list, Tx = 25)` tries to identify the "consistent correlation peak". `ccr.list` is the result of `lapply()` and `get_ccr_peaks(...)`, and `Tx` determines the size of a tolerance zone used in identifying the consistent correlation peak. `get_ccp(...)` returns `NULL` if there is no consistent correlation peak.

3.  **determine the congruent registration position:** A candidate position "receives" `votes` from basis segments that identify it or a close position within a tolerance zone of `Tx` as a candidate position in step 2.
    Votes for all candidate positions are tallied, and the position with the highest number of votes gets chosen as the *congruent registration position*, indicating that most of the basis segments find their highly similar counterpart in the reference signature in terms of correlation at this registration position.
    In case of ties the middle position is taken as the congruent registration position.
    Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS).
    The total number of CMPS is the CMPS score of the comparison.
    `get_CMPS(input.ccp, Tx = 25)` is the function that tallies the votes and determines the congruent registration position and congruent matching profile segments (CMPS).

Note that there are several parameters in the CMPS algorithm that are left to the users to decide, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks.set` identified on each level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3.
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters.
However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans.
Further research is needed, until then users are advised to think of default values as starting values and consider alternatives.
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`.
Here we present it with its default parameters.

@@

The function `extract_feature_cmps` allows for the following input from users besides the previously discussed parameters `seg_length`, `npeaks.set`, and `Tx`:

-   `x` and `y` are two signatures: `x` serves as the comparison signature (which will be divided into basis segments) and `y` is the reference signature;

-   `include` determines the format of the function result.
    Besides the `CMPS.score`, other aspects of the comparison help in understanding how the `CMPS.score` is computed.
    By default `include` is set to `NULL` and only the `CMPS.score` is returned; further results are included when `include` is (an abbreviation of) one of or a vector of the following strings: `"nseg", "congruent.pos", "congruent.seg.idx", "segments", "parameters"`, and `"full_result"`.
    If `include` is specified as `"full_result"` (or its abbreviation), the output includes everything listed below.

    -   `nseg`: the number of basis segments from the comparison signature; this is also the highest possible CMPS score of the comparison;
    -   `congruent.pos`: the congruent registration position;
    -   `congruent.seg.idx`: the indices of all congruent matching profile segments;
    -   `ccp.list`: a list showing identified candidate positions of all basis segments;
    -   `pos.df`: a data frame containing all candidate positions and their respective number of votes;
    -   `segments`: a list containing all basis segments;
    -   `parameters`: a list containing all input arguments of `extract_feature_cmps`;

-   `outlength` specifies the segment length of a basis segment at each level under the multi-segment lengths strategy.
    By default `outlength` is set to `NULL`, indicating that a basis segment should double its segment length at the next level and conforming to the description in @cmps .

@@

**An Example**

@@

`bullets$sigs` is a list of twelve numeric vectors corresponding to the twelve bullet signatures shown in \autoref{fig:sigs}.
`bullets$source` contains the URLs to the corresponding x3p file containing the topographic scan from the NIST Ballistics Toolmark Research Database [@nistdb].

@@

The signatures of Land 4 of Bullet 1 and Land 5 of Bullet 2 are stored in objects `sigs2` and `sigs1`, respectively.
This comparison consists of a pair of signatures that are known to be a match -- a KM (known match) comparison.
We compute the CMPS score using two versions of the CMPS algorithm:

@@

In the first example, `npeaks.set` is a vector of three integers, i.e. the algorithm uses the multi-segment lengths strategy to create the result object `cmps_with_multi_scale`.
For `cmps_without_multi_scale` each basis segment is linked to the top 5 candidate positions.
We use `include = "full_result"` to capture all results.
In this example, the CMPS is 9 when using multiple segments, and 12 when using a single segment.
As discussed in @cmps, using multi-segment lengths strategy can reduce the number of false positives when identifying candidate positions; however, any score based method is walking the line between false positives and false negatives.
As the number of false positives is reduced the number of false negatives might rise.
More discussion and comparisons between the two versions of the CMPS algorithm will be presented in later sections.
Note, that the multi-segment lengths method is slower because the algorithm is run once for each segment length.

@@

**Visualize and Understand CMPS results**

We also implemented graphing tools for visualizing results of the CMPS algorithm.
The goal is to provide users with tools to inspect each of the basis segments and to help them have a better understanding of how the algorithm works.
\autoref{fig:sigplots} shows the plots generated by the first graphing function, `cmps_signature_plot()`, and continues with the example above.
`cmps_signature_plot()` takes the output of `extract_feature_cmps(..., include = "full_result")` and returns a list of 5 elements.
It creates an overall impression of how the comparison signature aligns with the reference signature at the congruent registration position.

-   The first element is a plot called `segment_shift_plot`, shown in \autoref{fig:sigplot}. On this plot the reference signature is drawn as a black line, congruent matching profile segments from the comparison signature are overlaid in red at the congruent registration position.

-   The second plot is called `signature_shift_plot`, shown in \autoref{fig:sigplot2}. This visual presents both the comparison signature and the reference signature. The comparison signature is aligned with the reference signature based on the congruent registration position. Congruent matching profile segments are highlighted by solid red lines.

```{=tex}
\begin{figure}[hbt]
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:sigplot}The black line shows the comparison signature; each red line segment shows one congruent matching profile segment.}
```{r sigplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
sig.plot <- cmps_signature_plot(
  cmps_with_multi_scale
)
sig.plot$segment_shift_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:sigplot2}The black line shows the reference signature; the red line shows the comparison signature. Solid part shows the congruent matching profile segments, and the dashed part shows segments that do not agree with the congruent registration position.}
```{r sigplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
sig.plot$signature_shift_plot
```
\end{subfigure}
\caption{\label{fig:sigplots} Two visual outputs of cmps\_signature\_plot()}
\end{figure}
```
-   Other elements of this list are `seg_shift` and `sig_shift`. `sig_shift` gives the congruent registration position, while `seg_shift` is a data frame showing the congruent matching profile segments and their identified candidate position closest to the congruent registration position.

@@

While `cmps_signature_plot()` focuses on the signature level, `cmps_segment_plot()` focuses on the segment level.
It provides the "full result" of `extract_feature_cmps()`, but also takes an argument, `seg.idx`, indicating which segment should be inspected.
When checking `sig.plot$seg_shift` we notice that segment number 6 is not one of the congruent matching profile segments.
We can therefore set `seg.idx = 6` in `cmps_segment_plot()` and investigate the reason why this segment disagrees with the congruent registration position.

For each segment scale, we have two plots: `segment_plot` and `scale_ccf_plot`, as shown in \autoref{fig:segplots} for the example of segment number 6:

-   \autoref{fig:segplot} is the `segment_plot` for basis segment 6 at level one (in its original length). We used `npeaks.set = c(5, 3, 1)` in `extract_feature_cmps()` when calculating the CMPS score. Therefore the top five peaks are identified in the cross-correlation curve at level one. Segment 6 is plotted at the positions where these 5 peaks are identified with dashed lines in the `segment_plot`. The solid thick black line shows the segment at its original position (which in this example is very close to the actual registration position).
-   \autoref{fig:segplot2} is the `scale_ccf_plot` of basis segment 6 at level one. It shows the cross-correlation curve computed by the reference signature and the level-one basis segment 6. The five highest peaks are marked by dots on the curve.

```{=tex}
\begin{figure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot}segment\_plot for segment 6 at level one. The original position of segment 6 is indicated by the solid black line. Positions where the segment achieves the 5 highest cross-correlations are indicated by the dashed line segments.}
```{r segplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot <- cmps_segment_plot(
  cmps_with_multi_scale, 
  seg.idx = 6
)
seg.plot[[1]]$segment_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot2}scale\_ccf\_plot shows the cross-correlation curve between the reference signature and segment 6 at level one. The five highest peaks are marked by dots. The vertical red dashed line indicates the congruent registration position; the green dashed line shows a peak position in the highest segment level; the blue dashed lines show the tolerance zone around the green dashed line. We can see that none of the five highest peaks at level one falls within the tolerance zone, indicating that there is no consistent correlation peak or a candidate position identified by basis segment 6 under the multi-segment lengths strategy. Thus, the basis segment 6 doesn't vote for the congruent registration position and is not a cmps.}
```{r segplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot[[1]]$scale_ccf_plot
```
\end{subfigure}
\caption{\label{fig:segplots} Two plots used to investigate basis segment 6 at level one}
\end{figure}
```
Additionally, users can have more insights about why segment 6 is not a congruent matching profile segment if we put the `segment_plot` and `scale_ccf_plot` of all three segment levels together, as shown in \autoref{fig:seg_all} with the help of `ggpubr::ggarrange()`.

@@

## Evaluation Metrics

**Metrics Based on CMPS scores**

@@

**Scaled CMPS Scores**

@@

**Sum of Squares Ratio**

@@

**Comparing CMPS metrics and Parameters**

@@

**Comparing with the Original Results and the Random Forest model**

>>>>

---
title: An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets
header-includes:
  - \usepackage{amsmath}
  - \usepackage{xcolor}
  - \usepackage{longtable}
  - \usepackage{adjustbox}            
  - \usepackage[font=small,skip=5pt]{caption}
  - \usepackage{subcaption}
  - \usepackage{afterpage}
  - \usepackage{bbm}
author:
  # see ?rjournal_article for more information
  - name: Wangqian W. Ju
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
    url: https://github.com/willju-wangqian
    email:  wju@iastate.edu
    orcid: 0000-0002-9977-377X
  - name: Heike Hofmann
    url: https://github.com/heike
    email: hofmann@iastate.edu
    orcid: 0000-0002-9079-593X
    affiliation: Department of Statistics
    address:
    - Center for Statistics and Applications in Forensic Evidence
    - Iowa State University
    - 2438 Osborn Dr
    - Ames, IA 50011
thanks: |
  This work was partially funded by the Center for Statistics and Applications in Forensic Evidence (CSAFE) through Cooperative Agreement 70NANB20H019 between NIST and Iowa State University, which includes activities carried out at Carnegie Mellon University, Duke University, University of California Irvine, University of Virginia, West Virginia University, University of Pennsylvania, Swarthmore College and University of Nebraska, Lincoln.    
abstract: >
  In this paper, we introduce the R package \pkg{cmpsR}, an open-source implementation of the Congruent Matching Profile Segments (CMPS) method developed at the National Institute of Standards and Technology (NIST) for objective comparison of striated tool marks. The functionality of the package is showcased by examples of bullet signatures that come with the package. Graphing tools are implemented in the package as well for users to assess and understand the CMPS results. Initial tests were performed on bullet signatures generated from two sets of 3D scans in the Hamby study under the framework suggested by the R package \pkg{bulletxtrctr}. New metrics based on CMPS scores are introduced and compared with existing metrics. A measure called Sum of squares ratio is included, and how it can be used for evaluating different scans, metrics, or parameters is showcased with the Hamby study data sets. An open-source implementation of the CMPS algorithm makes the algorithm more accessible, generates reproducible results, and facilitates further studies of the algorithm such as methods comparison.
preamble: |
  % Any extra LaTeX you need in the preamble

@@

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "\\textwidth",
  fig.align = "center",
#  cache=TRUE,
  dpi = 100,
  warning = FALSE,
  message = FALSE
)
```

@@

**Hamby data set**
<!-- ### Hamby Data Set -->

@@

The automated framework used in this paper was proposed by @aoas.
In order to obtain bullet signatures from the x3p files of bullet land engravings, we use the `R` packages \pkg{x3ptools} [@x3ptools] and \pkg{bulletxtrctr} [@bulletxtrctr].
\pkg{x3ptools} is a package to read, write, and generally, process x3p files.
The \pkg{bulletxtrctr} package implements a pipeline for extracting and comparing signatures from scans of land-engraved areas.

@@

The white horizontal line in \autoref{fig:process}(a) indicates which cross-section was identified by the algorithm to represent the LEA; \autoref{fig:process}(b) shows the corresponding cross-sectional view.
Groove engraved areas are removed from the analysis as indicated by the vertical blue lines in \autoref{fig:process}(c) and \autoref{fig:process}(d)).
A non-parametric LOESS smooth [@loess] is fitted to capture the bullet curvature (\autoref{fig:process}(e)) and, finally, the **bullet signature** (\autoref{fig:process}(f)) is obtained as residuals of the cross-section and the fitted smooth.
Note, that in @cmps bullet signatures are referred to as bullet profiles.
However, to avoid confusion, we distinguish the notion of bullet signatures from bullet profiles.
Bullet profiles are shown in panels (b), (c), and (d) of \autoref{fig:process}, while \autoref{fig:process}(f) shows the corresponding bullet signature.
Identifying the groove engraved areas correctly is fundamental for a correct down-stream analysis of the signatures.
To allow for a human-in-the middle inspection and intervention we have provided an interactive web-application, implemented as `R` Shiny App [@shiny], \pkg{bulletinspectR} to identify and correct those errors.
An example of the extraction process with corresponding code and parameter settings can be found on Github at \url{https://github.com/willju-wangqian/CMPSpaper/tree/main/reproducible}. Note that the process of extracting signatures might be different from the one used in @cmps because no code or parameter settings are made available publicly.

@@

In this example, the comparison between \code{sig1} and \code{sig2}, two signatures coming from the same source (a known-match comparison), gets a CMPS score of 17; the comparison between \code{sig1} and \code{sig3}, two signatures coming from different sources (a known non-match comparison), gets a CMPS score of 1.

@@

1.  **cut the comparison signature into consecutive, non-overlapping and equal-length basis segments:** The command \code{get\_segs(x, len=50)} implements this step: it takes bullet signature \code{x} in the format of a numeric vector and cuts it into consecutive, non-overlapping and equal-length segments of length \code{len}, which are referred to as "basis segments".

@@

    -   **Calculate the cross-correlation curve:** Calculate the cross-correlation curve between a basis segment \code{x} and the reference signature \code{y} using the function \code{get\_ccf4(x, y, ...)} as shown in \autoref{fig:segplot2}. The position indicates the lag by which a basis segment is moved with respect to its original placement. A position is considered "good" if it results in a peak in the cross-correlation between the basis segment and the reference.

    -   **Correlation peaks:** Two strategies referred to as "multi-peak inspection" and "multi-peak inspection at different segment lengths" in @cmps are used for identifying positions of correlation peaks as candidate positions. The latter is also called "multi-segment lengths strategy". The parameter \code{npeaks.set} in \code{extract\_feature\_cmps(...)} determines which strategy to use and the number of candidate positions:

        -   If \code{npeaks.set} is an integer vector of length 1, for example \code{npeaks.set = 5}, the positions of the top five peaks in the cross-correlation curve are identified as candidate positions for registration.
        -   If \code{npeaks.set} is an integer vector of length more than 1, for example \code{npeaks.set = c(5, 3, 1)}, the multi-segment lengths strategy will be used: calculate the cross-correlation function between a basis segment and the reference and identify positions of the top five peaks; double the segment length to a specified value, re-calculate the cross-correlation function, and identify three peaks; repeat this process and identify a single peak in the newly computed cross-correlation function. \autoref{fig:seg_all} shows an example of three levels of basis segment 6 and their corresponding cross-correlation curves and identified peaks. Note that in @cmps the segment length is doubled at each level of a basis segment, but in the present implementation users are allowed to choose the segment length at each level.
        -   \code{get\_ccr\_peaks(comp, segments, seg\_outlength, nseg = 1, npeaks = 5)} computes the cross-correlation curve between a basis or increased segment and the reference signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to \code{npeaks}, which is an integer. \code{segments}, \code{seg\_outlength}, and \code{nseg} determines the segment in the cross-correlation computation, and \code{comp} gives the reference signature. If the multi-segment lengths strategy is used, then \code{get\_ccr\_peaks(...)} is called in a \code{lapply()} for each level of the basis segment. The resulting list is called \code{ccr.list}.

    -   **multi-segment lengths strategy:** with the multi-segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by \code{Tx} in all segment levels. Note that in @cmps, a segment at its largest scale (highest level) always identifies one peak, but we do not have this requirement in our implementation.

        -   the function \code{get\_seg\_scale(segments, nseg, out\_length)} is used to obtain the (potentially increased) version of a basis segment. \code{segments}, which is a list containing all basis segments generated by the function \code{get\_segs(...)} in step 1, and \code{nseg} are used to determine the basis segment to be increased. \code{out\_length} specifies the length of the output segment.
        -   \code{get\_ccp(ccr.list, Tx = 25)} tries to identify the "consistent correlation peak". \code{ccr.list} is the result of \code{lapply()} and \code{get\_ccr\_peaks(...)}, and \code{Tx} determines the size of a tolerance zone used in identifying the consistent correlation peak. \code{get\_ccp(...)} returns \code{NULL} if there is no consistent correlation peak.

3.  **determine the congruent registration position:** A candidate position "receives" votes from basis segments that identify it or a close position within a tolerance zone of \code{Tx} as a candidate position in step 2.
    Votes for all candidate positions are tallied, and the position with the highest number of votes gets chosen as the *congruent registration position*, indicating that most of the basis segments find their highly similar counterpart in the reference signature in terms of correlation at this registration position.
    In case of ties the middle position is taken as the congruent registration position.
    Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS).
    The total number of CMPS is the CMPS score of the comparison.
    \code{get\_CMPS(input.ccp, Tx = 25)} is the function that tallies the votes and determines the congruent registration position and congruent matching profile segments (CMPS).

Note that there are several parameters in the CMPS algorithm that are left to the users to decide, such as the length of basis segments \code{seg\_length} in step 1, the number of peaks \code{npeaks.set} identified on each level in step 2, and the length of the tolerance zone \code{Tx} in both step 2 and 3.
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters.
However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans.
Further research is needed, until then users are advised to think of default values as starting values and consider alternatives.
The main function that combines all steps in the CMPS algorithm described above is called \code{extract\_feature\_cmps(...)}.
Here we present it with its default parameters.

@@

The function \code{extract\_feature\_cmps} allows for the following input from users besides the previously discussed parameters \code{seg\_length}, \code{npeaks.set}, and \code{Tx}:

-   \code{x} and \code{y} are two signatures: \code{x} serves as the comparison signature (which will be divided into basis segments) and \code{y} is the reference signature;

-   \code{include} determines the format of the function result.
    Besides the \code{CMPS.score}, other aspects of the comparison help in understanding how the \code{CMPS.score} is computed.
    By default \code{include} is set to \code{NULL} and only the \code{CMPS.score} is returned; further results are included when \code{include} is (an abbreviation of) one of or a vector of the following strings: \code{"nseg"}, \code{"congruent.pos"}, \code{"congruent.seg.idx"}, \code{"segments"}, \code{"parameters"}, and \code{"full\_result"}.
    If \code{include} is specified as \code{"full\_result"} (or its abbreviation), the output includes everything listed below.

    -   \code{nseg}: the number of basis segments from the comparison signature; this is also the highest possible CMPS score of the comparison;
    -   \code{congruent.pos}: the congruent registration position;
    -   \code{congruent.seg.idx}: the indices of all congruent matching profile segments;
    -   \code{ccp.list}: a list showing identified candidate positions of all basis segments;
    -   \code{pos.df}: a data frame containing all candidate positions and their respective number of votes;
    -   \code{segments}: a list containing all basis segments;
    -   \code{parameters}: a list containing all input arguments of \code{extract\_feature\_cmps};

-   \code{outlength} specifies the segment length of a basis segment at each level under the multi-segment lengths strategy.
    By default \code{outlength} is set to \code{NULL}, indicating that a basis segment should double its segment length at the next level and conforming to the description in @cmps .

@@

**An example**

@@

\code{bullets\$sigs} is a list of twelve numeric vectors corresponding to the twelve bullet signatures shown in \autoref{fig:sigs}.
\code{bullets\$source} contains the URLs to the corresponding x3p file containing the topographic scan from the NIST Ballistics Toolmark Research Database [@nistdb].

@@

The signatures of Land 4 of Bullet 1 and Land 5 of Bullet 2 are stored in objects \code{sigs2} and \code{sigs1}, respectively.
This comparison consists of a pair of signatures that are known to be a match -- a KM (known match) comparison.
We compute the CMPS score using two versions of the CMPS algorithm:

@@

In the first example, \code{npeaks.set} is a vector of three integers, i.e. the algorithm uses the multi-segment lengths strategy to create the result object \code{cmps\_with\_multi\_scale}.
For \code{cmps\_without\_multi\_scale} each basis segment is linked to the top 5 candidate positions.
We use \code{include = "full\_result"} to capture all results.
In this example, the CMPS is 9 when using multiple segments, and 12 when using a single segment.
As discussed in @cmps, using multi-segment lengths strategy can reduce the number of false positives when identifying candidate positions; however, any score based method is walking the line between false positives and false negatives.
As the number of false positives is reduced the number of false negatives might rise.
More discussion and comparisons between the two versions of the CMPS algorithm will be presented in later sections.
Note, that the multi-segment lengths method is slower because the algorithm is run once for each segment length.

@@

**Visualize and understand CMPS results**

We also implemented graphing tools for visualizing results of the CMPS algorithm.
The goal is to provide users with tools to inspect each of the basis segments and to help them have a better understanding of how the algorithm works.
\autoref{fig:sigplots} shows the plots generated by the first graphing function, \code{cmps\_signature\_plot()}, and continues with the example above.
\code{cmps\_signature\_plot()} takes the output of \code{extract\_feature\_cmps(..., include = "full\_result")} and returns a list of 5 elements.
It creates an overall impression of how the comparison signature aligns with the reference signature at the congruent registration position.

-   The first element is a plot called \code{segment\_shift\_plot}, shown in \autoref{fig:sigplot}. On this plot the reference signature is drawn as a black line, congruent matching profile segments from the comparison signature are overlaid in red at the congruent registration position.

-   The second plot is called \code{signature\_shift\_plot}, shown in \autoref{fig:sigplot2}. This visual presents both the comparison signature and the reference signature. The comparison signature is aligned with the reference signature based on the congruent registration position. Congruent matching profile segments are highlighted by solid red lines.

```{=tex}
\begin{figure}[hbt]
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:sigplot}The black line shows the comparison signature; each red line segment shows one congruent matching profile segment.}
```{r sigplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
sig.plot <- cmps_signature_plot(
  cmps_with_multi_scale
)
sig.plot$segment_shift_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:sigplot2}The black line shows the reference signature; the red line shows the comparison signature. Solid part shows the congruent matching profile segments, and the dashed part shows segments that do not agree with the congruent registration position.}
```{r sigplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
sig.plot$signature_shift_plot
```
\end{subfigure}
\caption{\label{fig:sigplots} Two visual outputs of cmps\_signature\_plot()}
\end{figure}
```
-   Other elements of this list are \code{seg\_shift} and \code{sig\_shift}. \code{sig\_shift} gives the congruent registration position, while \code{seg\_shift} is a data frame showing the congruent matching profile segments and their identified candidate position closest to the congruent registration position.

@@

While \code{cmps\_signature\_plot()} focuses on the signature level, \code{cmps\_segment\_plot()} focuses on the segment level.
It provides the "full result" of \code{extract\_feature\_cmps()}, but also takes an argument, \code{seg.idx}, indicating which segment should be inspected.
When checking \code{sig.plot\$seg\_shift} we notice that segment number 6 is not one of the congruent matching profile segments.
We can therefore set \code{seg.idx = 6} in \code{cmps\_segment\_plot()} and investigate the reason why this segment disagrees with the congruent registration position.

For each segment scale, we have two plots: \code{segment\_plot} and \code{scale\_ccf\_plot}, as shown in \autoref{fig:segplots} for the example of segment number 6:

-   \autoref{fig:segplot} is the \code{segment\_plot} for basis segment 6 at level one (in its original length). We used \code{npeaks.set = c(5, 3, 1)} in \code{extract\_feature\_cmps()} when calculating the CMPS score. Therefore the top five peaks are identified in the cross-correlation curve at level one. Segment 6 is plotted at the positions where these 5 peaks are identified with dashed lines in the \code{segment\_plot}. The solid thick black line shows the segment at its original position (which in this example is very close to the actual registration position).
-   \autoref{fig:segplot2} is the \code{scale\_ccf\_plot} of basis segment 6 at level one. It shows the cross-correlation curve computed by the reference signature and the level-one basis segment 6. The five highest peaks are marked by dots on the curve.

```{=tex}
\begin{figure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot}segment\_plot for segment 6 at level one. The original position of segment 6 is indicated by the solid black line. Positions where the segment achieves the 5 highest cross-correlations are indicated by the dashed line segments.}
```{r segplot, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot <- cmps_segment_plot(
  cmps_with_multi_scale, 
  seg.idx = 6
)
seg.plot[[1]]$segment_plot
```
\end{subfigure}
\begin{subfigure}[t]{\textwidth}
\caption{\label{fig:segplot2}scale\_ccf\_plot shows the cross-correlation curve between the reference signature and segment 6 at level one. The five highest peaks are marked by dots. The vertical red dashed line indicates the congruent registration position; the green dashed line shows a peak position in the highest segment level; the blue dashed lines show the tolerance zone around the green dashed line. We can see that none of the five highest peaks at level one falls within the tolerance zone, indicating that there is no consistent correlation peak or a candidate position identified by basis segment 6 under the multi-segment lengths strategy. Thus, the basis segment 6 doesn't vote for the congruent registration position and is not a cmps.}
```{r segplot2, opts.label="codefig", echo=TRUE, out.width='\\textwidth', warning = FALSE}
seg.plot[[1]]$scale_ccf_plot
```
\end{subfigure}
\caption{\label{fig:segplots} Two plots used to investigate basis segment 6 at level one}
\end{figure}
```
Additionally, users can have more insights about why segment 6 is not a congruent matching profile segment if we put the \code{segment\_plot} and \code{scale\_ccf\_plot} of all three segment levels together, as shown in \autoref{fig:seg_all} with the help of \code{ggpubr::ggarrange()}.

@@

## Evaluation metrics

**Metrics based on CMPS scores**

@@

**Scaled CMPS scores**

@@

**Sum of squares ratio**

@@

**Comparing CMPS metrics and parameters**

@@

**Comparing with original results and the random forest model**
